Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-07-31 analysis/paper/paper.Rmd
adjust the paragraph
first editing, add comments to the file

Taiwan had been involved in regional exchange network of East Asia and Southeast Asia since the tenth century and became part of a global trade systmes during the early 17th century due to western European expansion that connected European and Asia [@Liu2017]. Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains unanswered. Indirect effects of colonialism is emphasized increasingly to discuss the colonial impact on the indigenous people in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts might cause significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed to emphasize the ability of active adoptions of foreign effects, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experience of the colonial encounter and explore the agency of indigenous people by examining how those goods were distributed and used when obtained [@Dietler2005; @Mullins2011; @Scaramelli2005].

Despite colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, little had been discussed in archaeology about the reaction of indigenous people to the European colonization. @Berrocal2018 study the long-term record of Heping Dao in northern Taiwan and explore the effects of European colonization on the local populations by focusing on the environmental change and introduction of diseases, with a little mention of indigenous social complexity. Northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region [@Andrade2007; @Kang2012]. Although the location is relatively peripheral, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into indigenous communities in Yilan through local regional exchange network. Among those trade goods, ornament such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Body adornment is usually viewed as a signal in indigenous social life to represent and reinforce their status through the body as the scene to display [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age were usually associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000].

In this paper, I explore the evidence from a site spanning from late Iron Age to historical period by addressing the question of whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems, and later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. As @Lape2003 points out based on case studies in island Southeast Asia, the study of multiple episodes of culture contact has potential to expand our understanding of the causes and effects of culture contact both locally and worldwide. Northeastern Taiwan provides a good example to explore the indirect impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact based on a long-term perspective. The variety and cultural context of trade ornaments will be reviewed, and their capacity to inform on how the distribution of ornaments reflect social changes will be examined. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network, and the consumption of ornaments present some spatial pattern that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan.

# Cultural context of ornaments in northeastern Taiwan, Yilan

Prior to European colonization, the indigenous people in northern Taiwan had already established their own exchange network that defined as "inter-insular trade" by @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan had increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes, and was intensified in the 17th century when the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Despite being on the periphery of major trade ports, Keelung and Tanshui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people since the prehistoric time, is a alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous people and another native group called Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous group. This form of exchange spread widely and even the Spanish soldiers learned to use agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149]. Father Jacinto Esquivel also recorded how indigenous viewed it as prestige goods in their culture. A indigenous male who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used as to solve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. A funeral process in indigenous culture was recorded about how agate beads used commonly in ritual contexts, where more agate beads, pottery and cloths were be placed into the grave of more influential people to indicate a family's prestige [@LiandWu2006, pp. 153].

Those documents indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans, that is believed to be introduced earlier from Chinese traders. In 1642, the Dutch VOC defeated the Spanish and take over the northern Taiwan. They introduced feudal symbolism in an attempt to control the indigenous communities by asking indigenous leaders to attend annual ceremony for feudal practices [@Andrade2007, ch. 9; @Kang2016, ch. 4]. More beads entered indigenous settlements through the gift-giving that the Dutch provided goods based on the demand of native people to secure indigenous alliances in the annual ceremony or on the way of traveling [@Kang2016, ch. 6]. Thus, I assume that the way of gift giving resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities within existing consumer contexts during this period.

The documents written by Chinese historians in Qing dynasty usually described the decorative purposes of the ornaments and some documents had similar texts that might be copied from previous ones without actual investigation. According to those records, indigenous people in Yilan had a custom to wear ornaments in ceremonial contexts to display their wealth and status [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production that usually possessed by rich people. Other people wore agate beads or glass beads on their head or neck to participating various ceremonies. In 1895, at the beginning of Japanese colonization, @Ino2002 conducted a field survey for several aboriginal groups and found that golden necklaces had not been used in Yilan that time, but elder people still used beads.

Recent ethnographic research of the Kavalan tribe describes that agate beads were used in divination practice, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate bead is valuable object that are usually passed down from mother to daughter as a heirloom in shaman families without knowing the exact origin. Were not been using, beads would be often hided at invisible place such as on the roof beams to make sure no one sees that. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination that is consistent with the observation in European colonial period. It was usually the female shamans conducting divination ritual that passed down from their ancestors. Despite most aborigines today don’t know the origin of agate beads, the value of beads is still be highlighted through oral history and their scarcity. Those historical documents from the early 17th century to the modern ethnography described how local indigenous people use ornaments in local cultural contexts that represents some social roles or status. However, compared to European contact period, there is less mention of beads in Chinese contact period and the description about ornaments is only limited to their dressing culture.

# Kiwulan in northeastern Taiwan

```{r, KWL-map, fig.cap = "Kiwulan location in Southeast Asia", out.width= "400px", echo=FALSE}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was a rescue archaeology projects that carried out from 2001 to 2004 in advance of construction. The total area includes eight open area sections that created 262 squares (4m by 4m) reaching 3,814 meter squares in total [@Chen2007]. Most ornaments were found in situ in the context while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures that indicates it was a settlement. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age that also called proto-historical period [@Liu2017], and historical period starting from the 17th century with European contact in Taiwan.

The earliest record of direct European contact with indigenous people in Yilan can be traced back to 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. The diagnostic artifacts of European contact found at Kiwulan including An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. Those artifacts coupled with the original radiocarbon dates were used to identify the pre-European and post-European periods. The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified both from Chinese official records and large amount of Chinese daily life porcelains found at Kiwulan [@Chen1963].

@@

```{r, KWL-sam-area, fig.cap = "Forty squares were sampled from the largest open area (in red with square ID number)", out.width= "70%", echo=FALSE}
# mapping sampling area
ggplot(sample_units) +
  geom_sf(data = AD_data, fill = NA) +
  geom_sf(color = 'red') +
  geom_sf_text(aes(label = `坑號`), size = 2.5, alpha = 0.9) +
  theme_minimal() +
  theme_void() 
```

The ornaments studied in this paper found from 40 adjacent square of the largest open area that was located in the middle part of excavation (Figure \@ref(fig:KWL-sam-area). They were sampled because those units were more stratigraphically intact with the least disturbance by modern constructions. The Upper Layer Culture component could be divided into six sub contexts spanning from 14 century to 19 century with a hundred year interval according to original radiocarbon dates, excavation depth, and types of diagnostic porcelain [@Hsieh2009; @Wang2011]. The contexts were assigned numbers from 1 to 6 where L1 corresponds to the youngest context and L6, the bottom layer, refers to the oldest context. To obtain a more reliable chronology for comparison between episodes of culture contact,I reevaluate those contexts for each sampled square by focusing on Anping jars and stonewares that are believed manufactured and distributed in Southeast Asia through European trade route during the early 17th century [@Ketel2011]. Those two categories of ceramic coupled with radiocarbon dates were used to identify the pre-European and post-European periods. Following the previous numbering principle, I separated three episodes of time including pre-European contact period(L5, L6), European contact period(L4), and Chinese contact period (L2 and L1).

@@

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major category. A: agate beads, B: bells, C: glass beads and golden beads, D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D categories were from original excavation report.", out.width= "400px", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

To explore whether body ornaments can reflect the colonial influences on indigenous populations, this paper examines the variety, frequency, and spatial distribution of trade ornaments across different time periods, including pre-European contact, European contact, and Chinese contact. Those three analytic units could represent three scales of exchange network from small to large, and three degree of contact from weak to intense. Because the appearance of ornaments is highly related to the original raw material, the type of raw material is the first criterion of classification that determined five major categories. In each category, ornaments were classified into several subtype according to their shapes when variations were present in each category. The variation was defined by the difference identified in at least two variables, for example dimensional variables such as length, or stylish variables such as patterns. In order to compare the general pattern of changes for subtypes, less frequent subtypes were left out from the analysis, but will be discussed separately in later section.

Ornaments from the post-hole area accounts for the majority of the amount in total. Post-hole area provides a clear contexts and basis for comparison to see if there is any spatial pattern of distribution. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts for `r sample_pro`% of the total number of ornaments found in the same open area section. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in \@ref(tab:tidy-data-for-table) in terms of three time episodes. The image for each subtype in major category shows in figure \@ref(fig:ornaments-image).

```{r, plot-counts-by-period, fig.cap = "Frequency of each category of ornament with a separate plot showing different time periods individually"}
# categories of ornaments by period
ornaments_count <-
  ornaments_period %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()

@@

# Results

## Frequency between time periods
Figure \@ref(fig:plot-counts-by-period) shows the comparison between different time periods that indicates a general pattern that most ornaments existed already before European contact and the frequencies reached a peak during the European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate bead, metal ring, and bell. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then the frequency decreased in the European contact period and even much less in the Chinese contact period.

@@

The distribution of frequency for subtypes in each major category are presented in \@ref(fig:plot-all-ornaments-count). The distribution shows that agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. Agate beads can be divided into seven subtypes, in which hexagonal shape is the most common subtype that appeared before European contact and increased significantly during the European contact and then declined in the Chinese contact. Small oval type shows similar frequency before and during European contact but not found in the Chinese contact period, while waxy oval bead only found during European contact period. Higher variety of agate beads were found in European contact compared to other time periods. Similarly, metal rings can be classified into eight subtypes, in which wide small ring is the most common metal ornaments, followed by wide large and thin large shape. The greater varieties for those two categories might indicate the multiple origins due to global trade network stimulated by the Europeans. In contrast, copper bell and glass bead have less variety, but glass bead has larger number due to they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bell is the large shape with a wide variety of human face as a motif, while the common shape for glass bead is bead with length less than 1cm that are identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of all class of ornament by time periods"}
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))
```

```{r, plot-spatial-pattern-each, fig.cap= "Spatial pattern for each class of ornament by time periods"}
# filter golden bead
ornaments_period_join_gold <-
  ornaments_period_join %>% 
  filter(Categories %in% 'Golden bead')

@@

par(mfrow = c(3, 1))
invisible(
map2(kernels_for_plotting, 
    titles,
    ~invisible(nndist_hypo_testing(.x, 
                         main = .y)))
)
```

>>>>

The direct impacts of European colonialism on indigenous communities in East Asia were much less pronounced than island Southeast Asia and Oceania. Throughout East Asia direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on Indigenous peoples in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local Indigenous societies caused significant transformations of Indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. The agency of Indigenous peoples in the colonial contexts has been broadly discussed to understand the negotiations between colonized and colonizer, and the resistance and accommodations of Indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand Indigenous experiences of colonial encounters and explore the agency of Indigenous peoples by examining how goods were distributed and used [@Dietler2005; @Mullins2011; @Scaramelli2005].

Taiwan provides an ideal case study of indirect effects of colonialism because it has archaeological evidence spanning its initial involvement in regional exchange networks of East Asia and Southeast Asia since the tenth century. It became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia [@Liu2017]. Despite colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of Indigenous responses to the European colonization. @Berrocal2018 study \the long-term record of Heping Dao\ in northern Taiwan and explore the effects of European colonization on the local populations by focusing on the environmental change and introduction of diseases, with a little mention of indigenous social complexity. Yilan in Northeastern Taiwan is an especially unique location for exploring indirect effects due to its physically isolated location. The rugged \XXX mountains to the west of the Yilan Plain that made it not easily accessible to Europeans and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. The sea provides a way through which Indigenous people in Yilan could interact with other groups from different regions, including other native Indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into Indigenous communities in Yilan through local regional exchange network. One of the most common types of trade goods in this region were personal ornaments such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments are a social signal of an individual’s status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000].

In this paper, I explore the evidence from Yilan spanning from late Iron Age to historical period to address the question of whether indirect colonial influences on Indigenous populations can be detected. I focus on  the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems, and \later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on Indigenous societies by comparing archaeological evidence from different episodes of culture contact [\cf.@Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine their potential of the ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new spatial patterns in the distributions of ornaments at Kiwulan that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan due to \???.

# Cultural context of ornaments in Yilan, northeastern Taiwan

Prior to European colonization, the indigenous people in northern Taiwan had already established their own exchange network that has been described as "inter-insular trade" by @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan was increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes. This was intensified in the 17th century with the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Due to its location on the periphery of major trade ports, \Keelung and \Tamsui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further \afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149]. Father Jacinto Esquivel also recorded how Indigenous people viewed agate pieces as prestige goods in their culture. A indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used as to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. An Indigenous funeral process was recorded, documenting the use of agate beads in ritual contexts, with more agate beads, pottery and cloths placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. Those historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

In 1642, the Dutch VOC defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the Indigenous communities by asking Indigenous leaders to attend annual ceremony for demonstrate their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of indigenous people to secure alliances in the annual ceremony or during traveling [@Kang2016, ch. 6]. Thus, I assume that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from the \Qing dynasty contain some notes on the purposes of ornaments from Yilan. According to those records, Indigenous people in Yilan had a custom to wear ornaments in ceremonial contexts to display their wealth and status [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, @Ino2002 conducted a field survey for several indigenous groups and found that golden necklaces were not used in Yilan at that time, but elderly people still used beads.

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called “subli”, by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as a heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. It was usually the female shamans conducting divination rituals that passed the beads down from their ancestors. Despite most Kavalan people today not knowing the specific origin of their agate beads, the value of beads is still highlighted through oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to European contact period, there are fewer mentions of beads in Chinese contact period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

# Excavations at Kiwulan in northeastern Taiwan

```{r, KWL-map, fig.cap = "Map showing the relative location of Kiwulan, Keelung, and Tamsui in northeastern Taiwan", out.width= "400px", echo=FALSE}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was a rescue archaeology projects that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 squares (4 m by 4 m) reaching 3,814 m2 in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of excavation (\Figure 2). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700 - 1200 AD) and a Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age, and historical period starting from the 17th century with European contact in Taiwan.

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with a hundred year intervals according to original radiocarbon dates, excavation depth, and types of diagnostic porcelain [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each samples square I separated excavation units into three episodes of time: pre-European contact period (L5, L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. \Q:The start of European contact Kiwalan is indicated in the archaeological record by the presence of An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. I used the presence and absence of An-ping jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods. European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. I recongnise this as the Chinese colonization period, which is archaeological visible at Kiwulan by \Chinese porcelains and bricks that brought by Chinsese communities [@Hsieh2009]. Also, the description of Chinese migration can be identified from Chinese official records [@Chen1963].

@@

```{r, KWL-sam-area, fig.cap = "Forty squares were sampled from the largest open area (in red with square ID number)", out.width= "70%", echo=FALSE}

# Need to say that this is a map of excavation areas at Kiwulan, because the reader cannot see the site name anywhere in this figure. Need to add in here that each square is 4 x 4 m, and say which way north is. Or add to the map and scale and north arrow.

# mapping sampling area
ggplot(sample_units) +
  geom_sf(data = AD_data, fill = NA) +
  geom_sf(color = 'red') +
  geom_sf_text(aes(label = `坑號`), size = 2.5, alpha = 0.9) +
  theme_minimal() +
  theme_void() 
```

# The ornaments

@@

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major category. A: agate beads, B: bells, C: glass beads and golden beads, D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D categories were from original excavation report (Chen 2007).", out.width= "400px", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

Ornaments from the post-hole area accounts for the \majority of the amount in total. Post-hole area provides a clear contexts and basis for comparison to see if there is any spatial pattern of distribution. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts \for `r sample_pro` % of the total number of ornaments found in the same open area section. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in \@ref(tab:tidy-data-for-table) in terms of three time episodes. The image for each subtype in each major category is shown in Figure \@ref(fig:ornaments-image).

<---include the number of ornaments from burials above--->

<---Something sounds wrong here. You say ‘majority’ but then later in the paragraph you say it’s 0.5% of the total number? How can 0.5% be the majority? --->

```{r, plot-counts-by-period, fig.cap = "Frequency of each category of ornament with a separate plot showing different time periods individually"}
### delete one plot

# categories of ornaments by period
ornaments_count <-
  ornaments_period %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()

@@

# Results of hypothesis testing

In this section I report the tests of the three hypotheses about indirect impacts of European colonization on the Indigenous community at Kiwulan. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European contact, and a lesser diversity after Chinese contact, I compare frequencies of ornament types between different time periods. To test the prediction that after European contact were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

## Frequency of ornaments between time periods
Figure \@ref(fig:plot-counts-by-period) shows the comparison of ornament type frequencies between different time periods at Kiwulan. It indicates a general pattern that most ornament types were present before European contact. The frequencies reached a peak during the European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate bead, metal ring, and bell. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then their frequency decreased in the European contact period and was even much less in the Chinese contact period.

@@

The distribution of frequency for subtypes in each major category are presented in Figure \@ref(fig:plot-altogether). The distribution shows that agate beads and metal rings have greater quantity and \variety of shapes compared to copper bell and glass beads during the European contact period. Agate beads can be divided into seven subtypes, in which hexagonal shape is the most common subtype that appeared before European contact and increased significantly during the European contact and then declined in the Chinese contact. The small oval type shows a similar frequency before and during European contact but is not found in the Chinese contact period. Waxy oval beads were only found during European contact period. A higher variety of agate beads were found in European contact compared to other time periods. Similarly, metal rings can be classified into eight subtypes, in which wide small ring is the most common metal ornaments, followed by wide large and thin large shape. The greater varieties for those two categories might indicate multiple origins due to global trade networks stimulated by the Europeans. In contrast, copper bell and glass bead have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

<---I think we need an addition bit of analysis in this section to show the relationship between diversity and sample size. We naturally expect diversity to increase as sample size increases. So we want to test and see if the changes in diversity that you describe here are only the natural effects of increase/decrease in sample size, or there are ‘real’ increases in diversity that cannot be explained by sample size changes. The peer reviewers will want to know the answers to this. --->

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of all class of ornament by time periods"}

#I don’t see any reference to this in the text. If you don’t reference a figure in the text, you need to delete it from the manuscript, Why not use a colour scale for frequency? I think the viridis scale would be perfect for this

ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))
```

```{r, plot-spatial-pattern-each, fig.cap= "Spatial pattern for each class of ornament by time periods"}

# Seems like some ornament types have <5 pieces. Probably not worth to plot those. You can write in the caption ‘only those types with n > 5 are shown here’

# filter golden bead
ornaments_period_join_gold <-
  ornaments_period_join %>% 
  filter(Categories %in% 'Golden bead')

@@

# make the blue lines thicker to be more visible
par(mfrow = c(3, 1))
invisible(
map2(kernels_for_plotting, 
    titles,
    ~invisible(nndist_hypo_testing(.x, 
                         main = .y)))
)
```
